Tidy data structure to support exploration and modeling of temporal data

### 2018-12-04 main.Rmd
restarted conclusion

@wickham2014tidy coined the term "tidy data", which is a rephrasing of the second and third normal forms in relational databases but in a way that makes more sense to data scientists. A suite of paired verbs such as `gather` and `spread` [@R-tidyr] were used to describe the reshape processing from messy data to tidy data. Tidy data, which standardized the mapping from the semantics of a dataset to its physical representation, serves remarkably as a fundamental unit of data analysis. R package **ggplot2** [@wickham2009ggplot2] pioneered developing the grammar of graphics based on tidy data, thus making it possible mapping from data space to visual elements. **dplyr** [@R-dplyr] generalised a coherent and consistent set of verbs to handle a wide range of data transformation tasks. @r4ds argues that 80% of data analysis tasks can be solved with tidy tools while the remaining 20% requires other tools. [R]

@@

We go through the whole exploration circle, and keep iterating and refining until data insights gained. The tsibble data structure substantially lubricates between these modules for time-based pipelines. The cohesive and coherent framework results in cleaner, clearer and more expressive code.

@@

All user-facing functions can be roughly organized into three major chunks according to their functionality: vector functions (1d), table verbs (2d), and window family. Each chunk is an independent module, but works interdependently. Vector functions in the package mostly deal with time. When collapsing a tsibble to less granular interval, these atomic functions can be combined with the `index_by()` table verb to accomplish this. A different function results in easily switching to aggregation of different time resolution. Since these functions are not exclusive to a tsibble, they can be used in a variety of applications in conjunction with other packages. On the other hand, these tsibble verbs can incorporate many third-party vector functions to step out of current tsibble zone. It is generally easier to trace back the errors users encounter from separating 1d and 2d functions.

@@

Time representation could be arbitrary, for example R's native `POSIXct` and `Date` for versatile date-times, nano time for nanosecond resolution [@R-nanotime], and pure numbers in simulations. Yet ordered factors can also be a source of time, such as month names from January to December and weekdays from Monday to Sunday. Tsibble supports an extensive range of index types from numerics to nano time, but there might be custom indices used for some occasions, for example school semesters. These academic terms vary from one institute to another within an academic year, which is defined differently from a calendar year. New index would be immediately recognized by the software once `index_valid()` defined, as long as it can be ordered from past to future. The interval regarding semesters is further outlined through `pull_interval()`. As a result, the rest software methods such as `has_gaps()`, `count_gaps()` and `fill_gaps()` will have instant support for data that contains this new index.

@@

A tidy representation of time series data, and data pipelines to facilitate data analysis flow have been proposed and discussed. It can be noted that tidy temporal data gains greater flexibility in keeping data richness, making data transformation and visualisation easily. A set of verbs provides a fluent and fluid pipeline to work with tidy time series data in various ways.

A new visualization framework for handling time series with nesting and crossing structure is under development. The proposed data structure proves a natural input (unit time series and parent-children structure) that can be piped into a visualization program. As long as the program understands the embedded structure, it will produce sensible layout and interactivity.

The ground of time series modelling or forecasting is left untouched in this paper. The future plan is to bridge the gap between tidy data and model building. Currently, it is required to casting to matrix from tidy data and therefore building a model. But time series models should be directly applied to tidy data as other wrangling tools do, without such an intermediate step. In particular, a univariate time series model, like ARIMA and Exponential Smoothing, can be applied to multiple time series independently. A tidy format to represent model summaries and forecasting objects will be developed and implemented in the future. Model summaries include coefficients, fitted values, and residuals; forecasting objects.

When involving multiple variables, one need to distinguish nesting and crossing data variables. "Nesting" refers to a scenario where a variable is nested within another when each category of the former variable co-occurs with only one category of the latter (for example, `country` nested within `continent`); on the other hand, "crossing" means that a variable is crossed with another when every category of one variable co-occurs with every category of the other (for example, `country` crosses with `gender`). Nesting is a special case of crossing. The former exhibits a hierarchical ordering with lower-level category (children) nested within higher-level category (parent) (more categories in lower level); whereas the ordering of the latter makes no difference. Whilst constructing tidy temporal data, users also need to specify the key that identifies the structure on the data. Rather than creating a separate object that contains the data structure, like what the **hts** does, we propose a syntax-based interface to directly deal with variables, so that it takes advantage of data semantics. The expression of a vertical bar (`|`) (can be interpreted probabilistically as the conditional distribution for its parent) is used to indicate nesting variables (reflect depths), while a comma (`,`) for crossing variables. @wilkinson2006grammar discussed the distinction between nesting and crossing with respect to facets in graphics, and he used `/` and `*` instead for nested and crossed expressions. A crossing structure displays all unique and possible combinations of crossed variables, including those not found in the data, in a graphical layout. However, tidy temporal data finds only the combinations that occur in the data and intentionally respects the structure of incomplete combinations.nclude future time path and distributions generating prediction intervals.

>>>>

@wickham2014tidy coined the term "tidy data", which is a rephrasing of the second and third normal forms in relational databases but in a way that makes more sense to data scientists. A suite of paired verbs such as `gather` and `spread` [@R-tidyr] were used to describe the reshape processing from messy data to tidy data. Tidy data, which standardized the mapping from the semantics of a dataset to its physical representation, serves remarkably as a fundamental unit of data analysis. R package **ggplot2** [@wickham2009ggplot2] pioneered developing the grammar of graphics based on tidy data, thus making it possible mapping from data space to visual elements. **dplyr** [@R-dplyr] generalised a coherent and consistent set of verbs to handle a wide range of data transformation tasks. @r4ds argues that 80% of data analysis tasks can be solved with tidy tools while the remaining 20% requires other tools. [R] The prevalence of grammar

@@

We go through the whole exploration circle, and keep iterating and refining until data insights gained. The tsibble data structure substantially lubricates between these modules for time-based pipelines. The cohesive and coherent framework results in more graceful and expressive code.

@@

All user-facing functions can be roughly organized into three major chunks according to their functionality: vector functions (1d), table verbs (2d), and window family. Each chunk is an independent module, but works interdependently. Vector functions in the package mostly deal with time. When collapsing a tsibble to less granular interval, these atomic functions can be combined with the `index_by()` table verb to accomplish this. A different function results in easily switching to aggregation of different time resolution. Since these functions are not exclusive to a tsibble, they can be used in a variety of applications in conjunction with other packages. On the other hand, these tsibble verbs can incorporate many third-party vector functions to step out of current tsibble zone. It is generally easier to trace back the errors users encounter from separating 1d and 2d functions. (lost in a web of functions)

@@

Time representation could be arbitrary, for example R's native `POSIXct` and `Date` for versatile date-times, nano time for nanosecond resolution [@R-nanotime], and pure numbers in simulations. Yet ordered factors can also be a source of time, such as month names from January to December and weekdays from Monday to Sunday. Tsibble supports an extensive range of index types from numerics to nano time, but there might be custom indices used for some occasions, for example school semesters. These academic terms vary from one institute to another within an academic year, which is defined differently from a calendar year. New index would be immediately recognized by the software upon defining `index_valid()`, as long as it can be ordered from past to future. The interval regarding semesters is further outlined through `pull_interval()`. As a result, the rest software methods such as `has_gaps()`, `count_gaps()` and `fill_gaps()` will have instant support for data that contains this new index.

@@

A new data abstraction representing temporal data named as "tsibble" has been proposed, spotlighting the "tidy data" principles brought to time domain. Tidy data begins to take shape in the state of time with the introduction of contextual semantics: index and key. Declared index provides direct support to time variable; variables that comprise the key defines study subjects over time. These semantics further determines unique data entries required for a valid tsibble. No matter how temporal data arrives, tsibble respects time index and keeps data richness. A tsibble frictionlessly pops into transformation, visualization, modelling and smoothly shifts amongst, allowing for rapid iterations in gaining data insights.
